Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-12 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM

%---------------------------------------------------------------------------------------------------%
%% LATEX DEFINITIONS
% Basic editing
\newcommand{\nick}[1]{{\color{red}#1}}
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
\newcommand{\here}{{\color{blue} here}}
% Basic math, derivatives
\newcommand{\td}[2]{\frac{d #1 }{d #2}}
\newcommand{\ttd}[2]{\frac{d^2 #1 }{{d #2}^2}}
\newcommand{\pd}[2]{ \frac{ \partial #1}{ \partial #2 } }
\newcommand{\ppd}[2]{ \frac{ \partial^2 #1}{ {\partial #2}^2 } }
\newcommand{\mppd}[3]{ \partial_{#2 #3} #1 }
% Basic math, vectors and other
%\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\bvec}[1]{{\mathbf{#1}}}
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\tavg}[1]{\langle #1 \rangle}
\newcommand{\norm}[1]{ \left\| #1 \right\| }
\newcommand{\ip}[1]{ \langle #1 \rangle }
\newcommand{\eps}{\varepsilon}
% For real and imaginary, could use \Re or \Im, or \mathcal{R}, or \text{Re}
\newcommand{\Pe}{\mathrm{Pe}}

@@

	These filtration processes, however, rely on an understand of complex transport phenomenon in porous media. Further, the action of continually flowing groundwater can alter medium properties through erosive and transportive processes. Thus, it becomes necessary to understand the transport of tracers through complex media whose properties are changing {\em dynamically} in response to the intervening fluid flow.

	This proposal aims to develop a deeper understanding of the physical processes by which groundwater flow can alter properties of a porous medium and to characterize the associated changes in transport, e.g.~anomalous dispersion. A recognized complexity of naturally occurring media is that their macroscopic properties, e.g.~permeability and mechanical strength, can be highly heterogeneous and anisotropic. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.

@@

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse can occur suddenly and catastrophically. It is likely that all of the mechanisms listed above --- erosion, dissolution, transport, compaction, and gravitational collapse --- contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity. Sinkhole formation has been explored with preliminary experiments run at the Geophysical Fluid Dynamics Institute at Florida State University \cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers.

\todo[inline]{I am not sure if we will keep this much detail in the paragraphs below on the sinkhole stuff. It depends on how much weight we give sinkholes in the body of the proposal. Instead, we could replace this discussion with more depth about the tracer transport and anomalous dispersion.}

% Stuff we could include
% McCurdy and Wang work
% Multiphase work

% REMOVED this paragraph because it emphasized sinkholes too much.
%To take one example (of particular relevance in the state of Florida) the formation of sinkholes can be triggered by certain human activities such as groundwater pumping or the weight of run-off storage ponds \cite{sandhu2018fate}. When collapse occurs below a storage pond, the underlying reservoir of pristine groundwater can be severely contaminated, thus surrendering a natural resource of great economic and societal value. Clearly, a deeper understanding of the mechanisms leading to sinkhole formation, and more generally that underly earth's natural filtration system, could reap enormous benefits in predicting and mitigating such environmental disasters. Other applications include understanding and predicting contamination by nuclear waste and by hydraulic fracturing.

@@

>>>>

%---------------------------------------------------------------------------------------------------%
%% LATEX DEFINITIONS
% Basic editing
\newcommand{\nick}[1]{{\color{red}#1}}
\newcommand{\tocite}{{\color{blue}(to cite) }}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
\newcommand{\here}{{\color{blue} here}}
% Basic math, derivatives
\newcommand{\td}[2]{\frac{d #1 }{d #2}}
\newcommand{\ttd}[2]{\frac{d^2 #1 }{{d #2}^2}}
\newcommand{\pd}[2]{ \frac{ \partial #1}{ \partial #2 } }
\newcommand{\ppd}[2]{ \frac{ \partial^2 #1}{ {\partial #2}^2 } }
\newcommand{\mppd}[3]{ \partial_{#2 #3} #1 }
% Basic math, vectors and other
%\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\bvec}[1]{{\mathbf{#1}}}
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\tavg}[1]{\langle #1 \rangle}
\newcommand{\norm}[1]{ \left\| #1 \right\| }
\newcommand{\ip}[1]{ \langle #1 \rangle }
\newcommand{\eps}{\varepsilon}
% For real and imaginary, could use \Re or \Im, or \mathcal{R}, or \text{Re}
\newcommand{\Pe}{\mathrm{Pe}}

@@

	These filtration processes, however, rely on an understanding of complex transport phenomenon in porous media. Further, the action of continually flowing groundwater can alter medium properties through erosive and transportive processes. Thus, it becomes necessary to understand the transport of tracers through complex media whose properties are changing {\em dynamically} in response to the intervening fluid flow.

	This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow can alter porous-media properties and to characterize the associated changes in transport within the evolving medium, e.g.~anomalous dispersion. A recognized complexity of naturally occurring media is that their macroscopic properties, e.g.~permeability and mechanical strength, can be highly heterogeneous and anisotropic. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.

@@

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse can occur suddenly and catastrophically. It is likely that all of the mechanisms listed above --- erosion, dissolution, transport, compaction, and gravitational collapse --- contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity. Sinkhole formation has been explored with preliminary experiments run at the Geophysical Fluid Dynamics Institute (GFDI) at Florida State University \cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of this project is to merge these two lines of inquiry, on erosive processes and sinkhole formation, to gain a deeper understanding the influence of mechanically or chemically induced erosion on gravitational collapse. Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for transport and dynamics of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models \tocite to be used for the quantitative description and prediction of sinkhole collapse.

	In the following, we first describe the governing equations underlying the proposed work, then detail the computational challenges for simulation erosion, characterizing transport, and potentially modeling gravitational interactions.

% Other stuff that could be included: McCurdy and Wang work, Multiphase work

@@

\vsp{3}
